---
title: 'WGCNA modules - filter on enrichment with post-mortem interval (PMI) or surgery injury DE genes'
author: "Jon Thompson, Pers lab"
date: "`r Sys.time()`" 
params: 
  date: !r substr(gsub("-","",as.character(Sys.Date())),3,1000)
  randomSeed: !r 12345
  pValThreshold: !r 0.05
  nrepBootstrap: !r 1e4
  padjMethod: "bonferroni"
output:
  html_notebook: 
  df_print: paged
number_sections: yes
toc: yes
toc_depth: 3
toc_float: yes
html_document:
  df_print: paged
---
  
Epilepsy temporal lobe 10x samples

## Setup

## Set options

```{r}
options(stringsAsFactors = F, 
        use="pairwise.complete.obs", 
        warn=1)
```

## Load packages

```{r, include=F}
library("dplyr")
library("Matrix")
library("parallel")
library("readr")
library("openxlsx")
library("data.table") 
library("here")

#TODO are these needed? 
if (F) {
  library("ggplot2")
  library("RColorBrewer") 
  library("Seurat")
  library("limma")
  library("lme4")
  library("gprofileR")
}
#"car","ComplexHeatmap", "liger", "WGCNA", "circlize", "xlsx"))
```

## source functions

```{r}
source(file=here("perslab-sc-library","utility_functions.R"))
source(file=here("perslab-sc-library","functions_sc.R"))
```

## constants 

```{r}
prefixData <- "ep_10x_v2"
prefixRun <-  "1"
prefixOut <- "ep_10x_v2"

colGeneNames = "genes"
colGeneWeights ="pkMs"
colModule = "module_merged"
colCellCluster = "cell_cluster_merged"
```

```{r}
params = list("randomSeed"=12345, 
              "date"=substr(gsub("-","",as.character(Sys.Date())),3,1000), 
              "nrepBootstrap"=1e4, 
              "pValThreshold" =0.05)
padjMethod = "bonferroni"
set.seed(params$randomSeed)
```

## load data

Load modules dataframe

```{r}
path_dt_geneMod = here("tables",paste0(prefixData, "_", prefixRun, "_geneMod_merged.csv.gz"))
dt_geneMod <- fread(path_dt_geneMod)
head(dt_geneMod)
```

Load embeddings data.table

```{r}
path_dt_embed = here("tables", paste0(prefixData, "_", prefixRun, "_kIM_cellModEmbed.csv.gz"))
dt_embed <- fread(path_dt_embed)
dt_embed[0:5,0:5]
```

```{r}
pathMetadata <- here("data",paste0(prefixData,"_cell_sample_annot_filter.csv"))
dt_metadata <- fread(pathMetadata)
dt_metadata
```

## Filter modules

### Filter out redundant modules(pruned using gene_module_merge2.R)

```{r}
vec_modsToKeep <- dt_geneMod$module_merged %>% unique 
vec_modsToKeep <- vec_modsToKeep[!is.na(vec_modsToKeep) & nchar(vec_modsToKeep)>0]
```

filter embeddings and convert to matrix (removing cell_cluster and data columns)

```{r}
vec_modsToKeep
```

```{r}
length(vec_modsToKeep)
```

```{r}
colnames(dt_embed) <- gsub(".*__","", colnames(dt_embed))
dt_embed %>% dplyr::select(.data = ., c("cell_id", vec_modsToKeep)) -> dt_embed_filter
dim(dt_embed_filter)
```

```{r}
dt_embed_filter[0:3,0:5]
```

## Compute enrichment with surgery and post-mortem DE genes using an empirical dot product test

Load surgery and post-morten DE genes

```{r}
df_surgeryGenes <- read.xlsx(xlsxFile = here("data", "surgery_postmorten_DEgenes.xlsx"), sheet = "Neurosurgical\ DE\ genes", startRow = 1, colNames = T, rowNames = F)
head(df_surgeryGenes)
```

```{r}
df_PMIGenes <- read.xlsx(xlsxFile = here("data","surgery_postmorten_DEgenes.xlsx"), sheet = "Post-mortem\ DE\ genes", startRow = 1, colNames = T, rowNames = F)
head(df_PMIGenes)
```

Make vectors of -log10 p-values for surgery and PMI genes

We use p values because we don't care about the direction of the LFC

```{r}
vec_surgeryMinLog10Pval <- df_surgeryGenes$P.Value %>% log10(.) %>% '-'(.)
vec_surgeryLFCweighted <- vec_surgeryMinLog10Pval * df_surgeryGenes$log2FC
names(vec_surgeryLFCweighted) <- df_surgeryGenes$gene 
head(vec_surgeryLFCweighted)
tail(vec_surgeryLFCweighted)
```

```{r}
vec_PMIMinLog10Pval <- df_PMIGenes$P.Value %>% log10(.) %>% '-'(.)
vec_PMI_LFCweighted <- vec_PMIMinLog10Pval * df_PMIGenes$log2FC
names(vec_PMI_LFCweighted) <- df_PMIGenes$gene 
head(vec_PMI_LFCweighted)
tail(vec_PMI_LFCweighted)
```

Use all genes in experiment as background 

```{r}
path_vecGeneBackground <- here("data", paste0(prefixData, "_genesBackground.RDS.gz"))

if (!file.exists(path_vecGeneBackground)) {
  pathDatExpr <- here("data", "ep_10x_v2_logcounts.csv.gz")
  dt_datExpr <- fread(pathDatExpr)
  vec_genesBackground <- dt_datExpr$gene
  saveRDS(vec_genesBackground, file = path_vecGeneBackground, compress="gzip")
} else {
  vec_genesBackground <- readRDS(path_vecGeneBackground)
}
```

make random vectors of -log10*LFC values for surgery and PMI DE genes, respectively

```{r}
list_vec_surgeryLFCweighted_rand <- lapply(1:params$nrepBootstrap, function(i) {
  vec_surgeryLFCweighted_rand = vec_surgeryLFCweighted
  names(vec_surgeryLFCweighted_rand) <- sample(x = vec_genesBackground, size=length(vec_surgeryLFCweighted_rand),replace = F)
  vec_surgeryLFCweighted_rand
})
```

```{r}
list_vec_PMI_LFCweighted_rand <- lapply(1:params$nrepBootstrap, function(i) {
  vec_PMI_LFCweighted_rand <- vec_PMI_LFCweighted
  names(vec_PMI_LFCweighted_rand) <- sample(x = vec_genesBackground, size=length(vec_PMI_LFCweighted_rand),replace = F)
  vec_PMI_LFCweighted_rand
})
```

```{r}
vec_mods = unique(dt_geneMod[[colModule]])
vec_mods <- vec_mods[!is.na(vec_mods)]
length(vec_mods)
```

get module gene weights, i.e. kMs

```{r}
list_vec_kMs <- lapply(vec_mods, function(module) {
  vec_kMs <- dt_geneMod[[colGeneWeights]][dt_geneMod[[colModule]]==module]
  names(vec_kMs) <- dt_geneMod[[colGeneNames]][dt_geneMod[[colModule]]==module] 
  vec_kMs <- vec_kMs[!is.na(vec_kMs)]
  vec_kMs
})
names(list_vec_kMs) <- vec_mods
lapply(list_vec_kMs[1:3], head)
```

define the intersect dot product function

```{r}
fun_dotProd = function(vec_kMs, vec_LFCweighted) {
  vec_genesShared <- intersect(names(vec_kMs), names(vec_LFCweighted))
  if (length(vec_genesShared)==0) return(0)
  vec_kMs[match(vec_genesShared,names(vec_kMs))] %*% vec_LFCweighted[match(vec_genesShared,names(vec_LFCweighted))]
}
```

compute intersect dot products for surgery and PMI -log10 pval vectors

surgery injury

```{r}
vec_dotprodSurgery <- sapply(list_vec_kMs, function(vec_kMs) fun_dotProd(vec_kMs = vec_kMs, vec_LFCweighted=vec_surgeryLFCweighted))
head(vec_dotprodSurgery)
```

```{r}
vec_dotprodPMI <- sapply(list_vec_kMs, function(vec_kMs) fun_dotProd(vec_kMs = vec_kMs, vec_LFCweighted=vec_PMI_LFCweighted))
head(vec_dotprodPMI)
```

## Compute empirical p-values

Make a null distribution for each module - geneset pair

surgery / injury genes

```{r}
mat_dotprodSurgery_NULL <- safeParallel(fun=function(vec_LFCweighted_rand) {
                                              sapply(list_vec_kMs, function(vec_kMs) {
                                                fun_dotProd(vec_kMs=vec_kMs, vec_LFCweighted=vec_LFCweighted_rand)
                                              })
                                            },
                                        list_iterable = list("X"=list_vec_surgeryLFCweighted_rand),
                                        outfile = here("log", paste0(prefixOut, "_log_pardotprodtest_surgery_NULL.txt")), 
                                        simplify=T)
```

PMI

```{r}
mat_dotprodPMI_NULL <- safeParallel(fun=function(vec_LFCweighted_rand) {
                                        sapply(list_vec_kMs, function(vec_kMs) {
                                          fun_dotProd(vec_kMs=vec_kMs, vec_LFCweighted=vec_LFCweighted_rand)
                                        })
                                      }, 
                                    list_iterable = list("X"=list_vec_PMI_LFCweighted_rand),
                                    outfile = here("log", paste0(prefixOut, "_log_par_dotprodtest_PMI_NULL.txt")), 
                                    simplify=T)

```

```{r}
mat_dotprodPMI_NULL[0:5,0:5]
```

Plot NULL distributions

```{r}
hist(mat_dotprodSurgery_NULL[1,], breaks=100)
```

```{r}
hist(mat_dotprodPMI_NULL[1,], breaks=100)
```

compute empirical p-values

"greater"

surgery

```{r}
vec_dotprodSurgeryPvalEmp <- sapply(vec_mods, function(module) {
  (sum(mat_dotprodSurgery_NULL[module,]>=vec_dotprodSurgery[module])+1)/(length(mat_dotprodSurgery_NULL[module,])+1)
})
head(vec_dotprodSurgeryPvalEmp)
```

PMI 

```{r}
vec_dotprodPMIPvalEmp <- sapply(vec_mods, function(module) {
(sum(mat_dotprodPMI_NULL[module,]>=vec_dotprodPMI[module])+1)/(length(mat_dotprodPMI_NULL[module,])+1)
})
head(vec_dotprodPMIPvalEmp)
```

Which modules are significantly enriched for surgery or post-morten interval DE genes?
(correct for multiple testing)

surgery 

```{r}
vec_modsSurgery <-names(vec_dotprodSurgeryPvalEmp)[p.adjust(p = vec_dotprodSurgeryPvalEmp, method=padjMethod) <= params$pValThreshold]
vec_modsSurgery
```

PMI 

```{r}
vec_modsPMI <- names(vec_dotprodPMIPvalEmp)[p.adjust(p=vec_dotprodPMIPvalEmp,method = padjMethod) <= params$pValThreshold]
vec_modsPMI
```

Remove PMI or surgery enriched modules

```{r}
vec_modsToKeep <- vec_modsToKeep[!vec_modsToKeep %in% c(vec_modsPMI,vec_modsSurgery)]
```

```{r}
length(vec_modsToKeep)
```

## write out results

```{r}
dt_geneMod$module_filter_PMI_surgery <- dt_geneMod$module_merged
dt_geneMod$module_filter_PMI_surgery[!dt_geneMod$module_merged %in% vec_modsToKeep] <- NA
```

```{r}
dt_geneMod$cell_cluster_filter_PMI_surgery <- dt_geneMod$cell_cluster_merged
dt_geneMod$cell_cluster_filter_PMI_surgery[!dt_geneMod$module_merged %in% vec_modsToKeep] <- NA
```

```{r}
fwrite(dt_geneMod, file=here("tables","ep_10x_v2_1_geneMod_merged_filterPMIsurg.csv.gz"))
```
