---
title: 'Epilepsy 10x samples v2 - gene ontology analysis (using gProfiler)'
author: "Jon Thompson, Pers lab"
date: "`r Sys.time()`" 
params: 
  date: !r Sys.time()
  randomSeed: !r 12345
  pValThreshold: !r 0.05
  padjMethod: "bonferroni"
output:
  html_notebook: 
  df_print: paged
number_sections: yes
toc: yes
toc_depth: 3
toc_float: yes
html_document:
  df_print: paged
---

Sources
https://www.bioconductor.org/packages/release/bioc/vignettes/topGO/inst/doc/topGO.pdf
http://bioconductor.org/packages/release/BiocViews.html#___OrgDb
GOSemSim: https://bioconductor.org/packages/release/bioc/vignettes/GOSemSim/inst/doc/GOSemSim.html

## Setup

## Set options


```{r}
options(stringsAsFactors = F, 
        use="pairwise.complete.obs", 
        warn=1)
```
## source functions

```{r}
source(file=here("perslab-sc-library", "utility_functions.R"))
source(file=here("perslab-sc-library","functions_sc.R"))
```

## Load packages

```{r}

library("dplyr")
library("Matrix")
library("parallel")
library("readr")
library("openxlsx")
library("data.table")
library("here")
library("gprofiler2")
# ipak(c("dplyr", "ggplot2", "Matrix", "parallel", "RColorBrewer", "readr", "pheatmap", "GOSemSim", "AnnotationHub","org.Hs.eg.db", "GSEABase"))#,"ComplexHeatmap", "liger", "WGCNA", "circlize", "xlsx"))
```

### constants 

```{r}
prefixData <- "ep_10x_v2"
prefixRun <-  "1"
prefixOut <- "ep_10x_v2"

colGeneNames = "genes"
colGeneWeights ="pkMs"
colModule = "module_filter_PMI_surgery"
colCellCluster = "cell_cluster_filter_PMI_surgery"
```

```{r}
params = list("randomSeed"=12345, 
              "date"=substr(gsub("-","",as.character(Sys.Date())),3,1000), 
              "nrepBootstrap"=1e4, 
              "pValThreshold" =0.05)
padjMethod ="bonferroni"
set.seed(params$randomSeed)
nrepBootstrap = 10000
```

## load data

Load modules dataframe

```{r}
dt_geneMod <- fread(here("tables","ep_10x_v2_1_geneMod_merged_filterPMIsurg.csv.gz"))
head(dt_geneMod)
```

load linear model epilepsy results

```{r}
dt_lm_epi_confint <- fread(here("tables", paste0(prefixData, "_dt_lm_epi_confint.csv")))
```

load epilepsy geneset enrichment results

```{r}
dt_wilcoxonVIFPvals <- fread(here("tables", paste0(prefixData, "_", prefixRun, "_dt_wilcox_geneset_test.csv")))
```

## Analysis

First extract a list of gene weights for genes that were actually clustered in the module

```{r}
vec_mods_m_h_signif <- dt_wilcoxonVIFPvals$module[p.adjust(p=dt_wilcoxonVIFPvals$m_h, method = padjMethod)<=params$pValThreshold]
```

```{r}
list_pkMs <- lapply(vec_mods_m_h_signif, function(module) {
  vecOut<-dt_geneMod[[colGeneWeights]][dt_geneMod[[colModule]]==module]
  names(vecOut) <- dt_geneMod[[colGeneNames]][dt_geneMod[[colModule]]==module]
  vecOut
})
names(list_pkMs) <- vec_mods_m_h_signif
head(list_pkMs[[1]])
```


run gprofiler

```{r}
list_gost <- gprofiler2::gost(query=lapply(list_pkMs, names),
                                 organism = "hsapiens", 
                                 ordered_query = T, 
                                 significant=T,
                                 exclude_iea = T, 
                                 correction_method="bonferroni",
                                 user_threshold=params$pValThreshold,
                                 custom_bg = load_obj(here("data", paste0(prefixData, "_genesBackground.RDS.gz"))),
                                 sources=c("GO:BP", "GO:CC","GO:MF"))
```

```{r}
head(list_gost$result)
```

```{r}
vec_mods_GOsignif <- unique(list_gost$result$query)
vec_mods_GOsignif
```

all 9 m_h enriched modules returned GO results

make a reduced gprofiler just for the publication as table

```{r}
dt_gprofiler_subtable <- setDT(list_gost$result[,c("query","term_name", "p_value")])
```

## write out outputs

```{r}
fwrite(list_gost$result, here("tables",paste0(prefixData,"_",prefixRun,"_GOgprofiler2_full.csv")))
```

```{r}
openxlsx::write.xlsx(x=list_gost$result, file = here("tables",paste0(prefixData,"_",prefixRun,"_GOgprofiler2_full.xlsx")))
```

```{r}
fwrite(dt_gprofiler_subtable, here("tables",paste0(prefixData,"_",prefixRun,"_GOgprofiler2_sub.csv")))
```

