---
title: 'WGCNA modules - filter on linear relationship with sample > epi; compute epilepsy linear model betas'
author: "Jon Thompson, Pers lab"
date: "`r Sys.time()`" 
params: 
  date: !r substr(gsub("-","",as.character(Sys.Date())),3,1000)
  randomSeed: !r 12345
  pValThreshold: !r 0.05
  nrepBootstrap: !r 1e4
output:
  html_notebook: 
  df_print: paged
number_sections: yes
toc: yes
toc_depth: 3
toc_float: yes
html_document:
  df_print: paged
---
  
Epilepsy temporal lobe 10x samples

## Setup

## Set options

```{r}
options(stringsAsFactors = F, 
        use="pairwise.complete.obs", 
        warn=1)

```

## Load packages

```{r, include=F}
library("dplyr")
library("Matrix")
library("parallel")
library("readr")
library("openxlsx")
library("data.table")
library("here")

#TODO are these needed? 
if (F) {
  library("ggplot2")
  library("RColorBrewer") 
  library("Seurat")
  library("limma")
  library("lme4")
  library("gprofileR")
}
#"car","ComplexHeatmap", "liger", "WGCNA", "circlize", "xlsx"))
```

## source functions

```{r}
source(file=here("perslab-sc-library","utility_functions.R"))
source(file=here("perslab-sc-library","functions_sc.R"))
```

## constants 

```{r}
prefixData <- "ep_10x_v2"
prefixRun <-  "1"
prefixOut <- "ep_10x_v2"

colGeneNames = "genes"
colGeneWeights ="pkMs"
colModule = "module_filter_PMI_surgery"
colCellCluster = "cell_cluster_filter_PMI_surgery"
```

```{r}
params = list("randomSeed"=12345, 
              "date"=substr(gsub("-","",as.character(Sys.Date())),3,1000), 
              "nrepBootstrap"=1e4, 
              "pValThreshold" =0.05)

padjMethod = "bonferroni"
set.seed(params$randomSeed)
```

## load data

Load modules dataframe

```{r}
dt_geneMod <- fread(here("tables","ep_10x_v2_1_geneMod_merged_filterPMIsurg.csv.gz"))
head(dt_geneMod)
```

get filtered modules as a vector 

```{r}
dt_geneMod[,.(module_filter_PMI_surgery)][!is.na(dt_geneMod$module_filter_PMI_surgery)] [nchar(dt_geneMod$module_filter_PMI_surgery)>0][[1]] %>% unique -> vec_mods_filter
```

get cell clusters as vector

```{r}
dt_geneMod[,.(cell_cluster_filter_PMI_surgery)][!is.na(dt_geneMod$cell_cluster_filter_PMI_surgery)] [nchar(dt_geneMod$cell_cluster_filter_PMI_surgery)>0][[1]] %>% unique -> vec_cellCluster_filter
```

Load expression data

```{r}
pathDatExpr <- here("data", paste0(prefixData,"_logcounts.csv.gz"))
dt_datExpr <- fread(pathDatExpr)
dt_datExpr[0:5,0:5]
```

load cell annotations

```{r}
path_dt_cell_annot <- here("data",paste0(prefixData,"_cell_sample_annot_filter.csv"))
dt_cell_annot <- fread(path_dt_cell_annot, data.table = T)
colnames(dt_cell_annot)[1] <- "cell_id"
head(dt_cell_annot)
```

load module embeddings matrices (per cell cluster annotation, l2)

```{r}
list_dt_embed <- lapply(FUN=function(cellClust) {
  fread(file = here("tables",paste0(prefixData,"_",prefixRun,"_", cellClust,"_cellModEmbed.csv")))
}, X = vec_cellCluster_filter)
```

Add "Name" i.e. sample, "l2" celltype and epilepy status columns

```{r}
list_dt_embed <- lapply(list_dt_embed, function(dt_embed) {
  vec_idxRow = match(dt_embed$cell_id,dt_cell_annot$cell_id)
  dt_embed[,Name:= dt_cell_annot$Name[vec_idxRow],]
  dt_embed[,l2:=dt_cell_annot$l2[vec_idxRow],]
  dt_embed[,Alias:=dt_cell_annot$Alias[vec_idxRow],] #we need this for computing logfoldchange later
  dt_embed[,epilepsy:=ifelse(grepl("EP",Alias),1,0),] #we need this for computing logfoldchange later
  dt_embed
})
```

collapse embeddings matrix by celltype / sample combination 

```{r}
list_dt_embed_avg <- lapply(list_dt_embed, function(dt_embed){
  vec_colnames_tmp <- colnames(dt_embed)[colnames(dt_embed) %in% vec_mods_filter]
 dt_embed[,lapply(.SD, mean), by= .(Name,l2), .SDcols=vec_colnames_tmp]})
```

add Alias, i.e. disease/ctr status to collapsed matrices
also age and sex

```{r}
list_dt_embed_avg <- lapply(list_dt_embed_avg, function(dt_embed_avg){
  vec_Alias = sapply(dt_embed_avg$Name, function(name) unique(dt_cell_annot$Alias[dt_cell_annot$Name==name]))
  vec_epilepsy <- ifelse(grepl("EP",vec_Alias),1,0)
  vec_Age = sapply(dt_embed_avg$Name, function(name) unique(dt_cell_annot$Age[dt_cell_annot$Name==name]))
  vec_Sex = sapply(dt_embed_avg$Name, function(name) unique(dt_cell_annot$Sex[dt_cell_annot$Name==name]))
  dt_embed_avg[,epilepsy:= vec_epilepsy,]
  dt_embed_avg[,Alias:= vec_Alias,]
  dt_embed_avg[,Age:= vec_Age,]
  dt_embed_avg[,Sex:= vec_Sex,]
  dt_embed_avg
})
```

make model matrix for linear regression: module ~ sample

```{r}
list_mat_model <- lapply(list_dt_embed_avg,function(dt_embed_avg){
  mat_model <- model.matrix(object = ~Name, data=as.data.frame(dt_embed_avg))
  colnames(mat_model) <- gsub("Name","",colnames(mat_model))
  mat_model <- data.table(mat_model, "Biopsy"=ifelse(dt_embed_avg$Name=="Biopsy",1,0))
})
```

```{r}
list_mat_model_full <- mapply(FUN= function(mat_model, dt_embed_avg) {
  data.table(mat_model, dt_embed_avg)
  }, 
  mat_model =list_mat_model, dt_embed_avg = list_dt_embed_avg, SIMPLIFY=F)
```

```{r}
names(list_mat_model_full) <- vec_cellCluster_filter
```

Compare R squared stats for sample and condition, within the module's subtype only

```{r}
# Compute R squared with sample as covariate
list_mat_lmRsquared_l2_Name <- lapply(names(list_mat_model_full), function(cellClust) {
  mat_model_full <- list_mat_model_full[[cellClust]] 
  vec_modules <- colnames(mat_model_full)[colnames(mat_model_full) %in% vec_mods_filter]
      
  mat_tmp <- sapply(vec_modules, function(module) {
        vec_tmp <- sapply(unique(mat_model_full[["Name"]]), function(sampleX) {
          regr <- lm(eval(parse(text=module)) ~ eval(parse(text=sampleX)), data = mat_model_full) 
          summary.lm(regr)$r.squared
          })
      names(vec_tmp) <- unique(mat_model_full[["Name"]])
      return(vec_tmp)
    })
})
```

```{r}
mat_lmRsquared_l2_Name <- Reduce(x=list_mat_lmRsquared_l2_Name,f=cbind)
```

Compute R squared with epilepsy status as covariate

```{r}
list_vec_lmRsquared_l2_epi <- sapply(names(list_mat_model_full), function(cellClust) {
  mat_model_full <- list_mat_model_full[[cellClust]] 
  vec_modules <- colnames(mat_model_full)[colnames(mat_model_full) %in% vec_mods_filter]
  vec_tmp <- sapply(vec_modules, function(module) {
    regr<-lm(eval(parse(text=module)) ~ epilepsy, data = mat_model_full) # linear model with only epilepsy epi
    summary.lm(regr)$r.squared})
  names(vec_tmp) = vec_modules
  return(vec_tmp)
})
vec_lmRsquared_l2_epi <- unlist(list_vec_lmRsquared_l2_epi , use.names = T)
names(vec_lmRsquared_l2_epi) <- gsub(".*\\.","",names(vec_lmRsquared_l2_epi))
```

# Identify modules whose expression is better predicted by epilepsy than by any individual sample within their celltype of origin

```{r}
vec_logical_epilepsy_vs_Name <- sapply(names(vec_lmRsquared_l2_epi), function(mod) {
  all(mat_lmRsquared_l2_Name[,mod]<vec_lmRsquared_l2_epi[mod]) # if for any samples the R^2 is larger than for condition, evaluates to F
},simplify = T)
```

```{r}
table(vec_logical_epilepsy_vs_Name)
```

This yields 38 modules more strongly explained by epilepsy than by any sample

```{r}
vec_mods_filter_lm <- vec_mods_filter[vec_logical_epilepsy_vs_Name]
vec_mods_filter_lm
```


### what celltypes are the modules from? 

```{r}
vec_cellClust_filter_lm <- sapply(vec_mods_filter_lm, function(eachMod) {
dt_geneMod[[colCellCluster]][dt_geneMod[[colModule]]==eachMod] %>% unique
})
names(vec_mods_filter_lm) = vec_cellClust_filter_lm
vec_mods_filter_lm
```


### Compute lm again for remaining modules, with confidence intervals and t-stat p-values

```{r}
list_mat_lm_epi_confint <- sapply(names(list_mat_model_full), function(cellClust) {
  mat_model_tmp <- list_mat_model_full[[cellClust]] 
  vec_modules <- colnames(mat_model_tmp)[colnames(mat_model_tmp) %in% vec_mods_filter_lm]
  sapply(vec_modules, function(module) {
    regr<-lm(eval(parse(text=module)) ~ epilepsy, data = mat_model_tmp) # linear model with only epilepsy condition
    betaEst <- coef(regr)["epilepsy"]
    pval <- summary(regr)$coefficients["epilepsy", "Pr(>|t|)"]
    vec_confint <- stats::confint(object=regr, parm="epilepsy", level=0.95)
    vec_out <- c(betaEst, vec_confint, pval)
    names(vec_out) = c("beta", "lower","upper", "p-val")
    return(vec_out)
  }) %>% t -> mat_tmp 
  rownames(mat_tmp) = vec_modules
  return(mat_tmp)
})

mat_lm_epi_confint <- Reduce(x=list_mat_lm_epi_confint, f=rbind)

head(mat_lm_epi_confint)
```

## check significance after multiple testing

```{r}
vec_logicalBetaPvalSignif <- p.adjust(mat_lm_epi_confint[,"p-val"], method=padjMethod)<params$pValThreshold
table(vec_logicalBetaPvalSignif)
```

## Compute log fold change of the means between conditions

collapse full cell-level embeddings by condition
```{r}
list_dt_embed_avg_epi <- lapply(list_dt_embed, function(dt_embed){
  vec_colnames_tmp <- colnames(dt_embed)[colnames(dt_embed) %in% vec_mods_filter_lm]
  dt_embed[,lapply(.SD, mean), by= .(epilepsy), .SDcols=vec_colnames_tmp]})
```

```{r}

list_vec_LFC <- lapply(list_dt_embed_avg_epi, function(dt_embed_avg_epi) {
  (as.numeric(dt_embed_avg_epi[2,2:ncol(dt_embed_avg_epi)]) /  # epilepsy row
    as.numeric(dt_embed_avg_epi[1,2:ncol(dt_embed_avg_epi)]))  %>% # control row
    log2 -> vec_LFC
  names(vec_LFC) <- colnames(dt_embed_avg_epi)[2:ncol(dt_embed_avg_epi)]
  return(vec_LFC)})

vec_LFC <- unlist(list_vec_LFC, use.names=T)
names(vec_LFC) <- gsub(".*\\.","",names(vec_LFC))
vec_LFC
```

## gather outputs in single data.table
```{r}
dt_lm_epi_confint <- data.table("module"=rownames(mat_lm_epi_confint),
                                "l2_cell_cluster"=vec_cellClust_filter_lm,
                                mat_lm_epi_confint)
dt_lm_epi_confint[,LFC:=vec_LFC]

head(dt_lm_epi_confint)
```

```{r}
fwrite(dt_lm_epi_confint, file=here("tables", paste0(prefixOut, "_dt_lm_epi_confint.csv")))
```

```{r}
openxlsx::write.xlsx(x=dt_lm_epi_confint, file=here("tables", paste0(prefixOut, "_dt_lm_epi_confint.xlsx")))
```
