---
title: 'Epilepsy - plotting'
author: "Jon Thompson, Pers lab"
date: "`r Sys.time()`" 
params: 
  date: !r substr(gsub("-","",as.character(Sys.Date())),3,1000)
  randomSeed: !r 12345
  pValThreshold: !r 0.05
  padjMethod: "bonferroni"
output:
  html_notebook: 
  df_print: paged
number_sections: yes
toc: yes
toc_depth: 3
toc_float: yes
html_document:
  df_print: paged
---

   
# Setup

## Set options

```{r}
options(stringsAsFactors = F, 
        use="pairwise.complete.obs", 
        warn=1)
```

## Source functions

## Load packages


```{r, include=F}
library("magrittr")
library("Matrix")
library("openxlsx")
library("data.table") 
library("here")
library("ggplot2")
library("RColorBrewer") 
library("pheatmap")
#library("scales")
```

## source functions

```{r}
source(file=here("perslab-sc-library","utility_functions.R"))
source(file=here("perslab-sc-library","functions_sc.R"))
```

## constants 

```{r}
prefixData <- "ep_10x_v2"
prefixRun <-  "1"
prefixOut <- "ep_10x_v2"

colGeneNames = "genes"
colGeneWeights ="pkMs"
colModule = "module_filter_PMI_surgery"
colCellCluster = "cell_cluster_filter_PMI_surgery"
```

```{r}
params = list("randomSeed"=12345, 
              "date"=substr(gsub("-","",as.character(Sys.Date())),3,1000), 
              "nrepBootstrap"=1e4, 
              "pValThreshold" =0.05)
padjMethod="bonferroni"
set.seed(params$randomSeed)
```

## load data

```{r}
dt_geneMod <- fread(here("tables","ep_10x_v2_1_geneMod_merged_filterPMIsurg.csv.gz"))
head(dt_geneMod)
```

load linear model epilepsy results

```{r}
dt_lm_epi_confint <- fread(here("tables", paste0(prefixData, "_dt_lm_epi_confint.csv")))
head(dt_lm_epi_confint)
```

load epilepsy geneset enrichment results

```{r}
dt_wilcoxonVIFPvals <- fread(here("tables", paste0(prefixData, "_", prefixRun, "_dt_wilcox_geneset_test.csv")))
head(dt_wilcoxonVIFPvals)
```

## Plotting

## 1. Gene set enrichment plot

```{r}
set.seed(2)
```

```{r}
vec_celltype <- dt_wilcoxonVIFPvals$cellCluster
vec_color = sample(x=unique(gsub("\\d","",colors())), size=length(unique(vec_celltype)), replace=F)
vec_colors_celltype = rep(x = vec_color,times = c(dt_wilcoxonVIFPvals[,.N,by="cellCluster"][[2]]))

orderToUse1 = 1:length(vec_celltype)
vec_celltype <- factor(x = vec_celltype)
vec_celltype <- reorder(x=vec_celltype, X = orderToUse1, order=T)

vec_module <- dt_wilcoxonVIFPvals$module 
orderToUse2 = 1:length(vec_module)
vec_module <- factor(x = vec_module)
vec_module <- reorder(x=vec_module, X = orderToUse2, order=T)

names(vec_module) <- vec_module # for coloring

dt_wilcoxonVIFPvals[["m_h"]] %>% log10 %>% '-'(.) -> vec_genesetMinlog10pval
names(vec_genesetMinlog10pval) <- dt_wilcoxonVIFPvals$module
  
# put everything in data.frame for plotting
df_data =data.frame("celltype"=vec_celltype, 
                    "module"=vec_module, 
                    "color"=vec_colors_celltype, 
                    "minLog10pval" = vec_genesetMinlog10pval)

# Bonferroni adjusted p value cutoff
params$pValThreshold %>% '/'(length(vec_genesetMinlog10pval)) %>% log10 %>% '-'(.) -> cutoff
```

plot 

```{r}

p <- 
  ggplot(data=df_data, aes(x=module, y=minLog10pval, fill=celltype)) +
  geom_bar(stat="identity", position=position_identity(), group=df_data$celltype) + 
  scale_fill_manual(values=vec_color) + # this is a kluge
  #scale_fill_manual(values=unique(df_data$celltype)) + # this is a kluge
  geom_hline(yintercept=cutoff, linetype="dashed", color = "red") +
  scale_y_continuous("-log10(p-value)", expand = c(0, 0)) +
  scale_x_discrete("module") +  
  # geom_text(aes(label=module),
  #         #angle=90,
  #         #vjust=-1,
  #         hjust = 2,
  #         color="black",
  #         angle=90,
  #        position = position_dodge(), 
  #        size=2.5) +
  theme(axis.text.x = element_text(angle=90,hjust=1),#element_blank(),
      axis.line.x = element_blank(),
      axis.ticks.x = element_blank(),
      panel.grid.major = element_blank(), 
      panel.grid.minor = element_blank())

saveMeta(savefnc=ggsave,plot = p,
         filename = here("plots", paste0(prefixOut, "_genesetEnrichmentBarplot", params$date,".pdf")), height=10, width =15)

```

Plot betas as heatmap column annotation under enrichment barplot

```{r}
vec_beta_modCelltype <- dt_lm_epi_confint[["beta"]][match(df_data$module, dt_lm_epi_confint$module)]

names(vec_beta_modCelltype) <- as.character(df_data$module)
```

```{r}
pheatmap::pheatmap(mat = matrix(data = vec_beta_modCelltype, nrow=1),
         color =  rev(colorRampPalette(brewer.pal(n=9, name="RdYlBu"))(100))[1:60],
         kmeans_k = NA,
         cluster_rows = F,
         cluster_cols = F,
         legend_breaks=c(-0.2,0,0.15),
         main=NA,#"Log2 fold change epilepsy versus control",
         filename= here("plots", paste0(prefixOut, "_beta_heatmap_", params$date,".pdf")), cellheight = 40, height=1.2, width =15)

```

## hubgenes table 

use all 34 modules associated with epilepsy status - not just those also enriched for m_h

```{r}
# vec_mods_m_h_signif <- dt_wilcoxonVIFPvals$module[p.adjust(dt_wilcoxonVIFPvals$m_h, method = params$padjMethod) <= params$pValThreshold]

```

```{r}
mat_hubgenes <- sapply(dt_wilcoxonVIFPvals$module, function(module) {
  dt_geneMod[[colGeneNames]][dt_geneMod[[colModule]]==module][1:10]
})
mat_hubgenes
```


```{r}
saveMeta(savefnc = write.xlsx, x=mat_hubgenes, file=here("tables", paste0(prefixOut, "_", params$date,"_mat_hubgenes.xlsx")))
```

