---
title: 'Epilepsy 10x samples - preprocess for WGCNA'
author: "Jon Thompson, Pers lab"
date: "`r Sys.time()`" 
output:
  html_notebook: 
  df_print: paged
number_sections: yes
toc: yes
toc_depth: 3
toc_float: yes
html_document:
  df_print: paged
---

# Setup

## reference

conos vignette https://github.com/hms-dbmi/conos/blob/master/vignettes/walkthrough.md

## Options

```{r}
options(stringsAsFactors = F, 
        use="pairwise.complete.obs", 
        warn=1, 
        verbose=F,
        mc.cores=20,# for parallel computation
        max.print= 100, width = 50
        ) 
```

## Load packages

```{r}
library("here")
library("dplyr")
library("Matrix") 
library("parallel")
library("Seurat")
library("data.table")
library("conos")
library("pagoda2") # conos dependency
```

## Set constants

```{r}
flagDate = substr(gsub("-","",as.character(Sys.Date())),3,1000)
dirProject ="/projects/jonatan/tmp-epilepsy/"
prefixData = "ep_10x_v2"
prefixRun = "1"
randomSeed = 12345
set.seed(seed = randomSeed)
npcs = 5
```

# download and load data

## download 

```{r}
url_conosObj <- "http://pklab.med.harvard.edu/viktor/data/epilepsy/con_filt_samples_10x.rds.gz"
url_cell_annotation <- "http://pklab.med.harvard.edu/viktor/data/epilepsy/cell_annotation.csv"


path_conosObj <- here("data", paste0(prefixData,"_conos.rds.gz"))
path_cell_annot <- here("data", paste0(prefixData, "_cell_annot.csv"))
path_metadata <- here("data",paste0(prefixData, "_sample_metadata.csv" ))
```

```{r}
if (!file.exists(path_conosObj)) download.file(url=url_conosObj, destfile = path_conosObj)
if (!file.exists(path_cell_annotation)) download.file(url=url_cell_annotation, destfile=path_cell_annotation)
```

## load data

```{r}
conosObj <- readRDS(path_conosObj)
```

```{r}
df_cell_annot <- fread(path_cell_annot, data.table = F)
rownames(df_cell_annot) <- df_cell_annot$cell
df_cell_annot$cell <- NULL
head(df_cell_annot)
```

```{r}
dt_meta <- fread(path_metadata)
head(dt_meta)
```

# Analysis

## QC

extract lognormalized counts from conos object

```{r}
lapply(conosObj$samples, '[[', ("counts")) %>% Reduce(f = rbind, x = .) %>% t -> spMat_logCounts
```

```{r}
spMat_logCounts[0:4,0:4]
```

```{r}
dim(spMat_logCounts)
```

```{r}
df_cell_annot %>% nrow
```

filter cell annotations to retain only those present in lognormalised expression 

```{r}
df_cell_annot <- df_cell_annot[rownames(df_cell_annot)%in%colnames(spMat_logCounts),]
```

### Merge cell and sample cell_annotations into per cell metadata

```{r}
gsub("_[ATCG]{16}-\\d","", rownames(df_cell_annot)) %>% gsub("_one", "", .) -> vec_tmpc_cellPrefixes
unique(vec_tmpc_cellPrefixes)
```

```{r}
dt_meta$Name %>% unique
```

```{r}
dt_meta$Name[!dt_meta$Name %in% vec_tmpc_cellPrefixes]
```

```{r}
df_cell_annot$Name = vec_tmpc_cellPrefixes
```

```{r}
df_cell_annot$Alias <- dt_meta$Alias[match(df_cell_annot$Name, dt_meta$Name)]
df_cell_annot$Age <- dt_meta$Age[match(df_cell_annot$Name, dt_meta$Name)]
df_cell_annot$Sex <- dt_meta$Sex[match(df_cell_annot$Name, dt_meta$Name)]
```

```{r}
table(df_cell_annot$Sex)
```

What are the identities at different levels of cell cluster annotations?

```{r}
table(df_cell_annot$l1)
```

```{r}
table(df_cell_annot$l2)
```

```{r}
table(df_cell_annot$l3)
```

```{r}
table(df_cell_annot$l4)
```

create seurat object for PCA 

```{r}
seuratObj <- CreateSeuratObject(counts = spMat_logCounts, 
                                meta.data = df_cell_annot,
                                assay = "RNA",
                                project = prefixData)

```

```{r}
Idents(seuratObj) <- seuratObj$Name
```

look at PCA plots

```{r}
seuratObj %>% 
  FindVariableFeatures(., nfeatures = 3000)  %>% 
  ScaleData(., block.size= 10000, min.cells.to.block = 15000) %>%
  RunPCA(object = ., features = VariableFeatures(.), npcs = npcs, seed.use = randomSeed, verbose=F) -> 
  seuratObj
```

```{r, fig.width=8}
DimPlot(seuratObj, reduction = "pca", dims = c(3,4), group.by="Name")
```

# check correlations between sample and 20 first PCs

```{r}
mat_mod <- model.matrix(data = data.frame("sample"=Idents(seuratObj)), object = formula(x=~sample_name))
```

```{r}
cor(x = seuratObj@reductions$pca@cell.embeddings[,1:20], y = mat_mod[,-1], method = "pearson") %>% data.table("PC"=rownames(.),.) -> dt_PCAcor
```

```{r}
hist(x=as.matrix(dt_PCAcor[,-1]), breaks = 100)
```

```{r}
apply(X = dt_PCAcor[,-1], MARGIN = 2, FUN = function(sample_col) hist(x=sample_col))
```

samples 3,4,5,9, and 10 have quite skewed distributions of correlations with PCs

However, since sample confounds condition we can't go correcting the data. Better to filter later.

## write out files

write to file

logCounts 

```{r}
file.out.data = here("data", paste0(prefixData, "_logcounts.csv.gz"))
if (!file.exists(file.out.data)) {
  lapply(conosObj$samples, function(conosSample) {conosSample[["counts"]] %>% as.matrix %>% t %>% as.data.table}) %>% Reduce(f = cbind, x = .) -> dt_logCounts_tmp
  dt_logCounts_tmp <- data.table("gene" = rownames(spMat_logCounts), dt_logCounts_tmp) 
  fwrite(x=dt_logCounts_tmp, file= file.out.data, nThread = 40, compress = "gzip",showProgress = T)
  rm(dt_logCounts_tmp)
}
```

combined cell annotation and sample metadata, filtered to retain only cells present in logCounts

```{r}
file.out.meta = here("data", paste0(prefixData, "_cell_sample_annot_filter.csv"))
```

```{r}
if (!file.exists(file.out.meta)) fwrite(x = df_cell_annot, file = file.out.meta, nThread=20, showProgress = T, row.names = T)
```
